---
title: "gaussian mixture models"
author: "Nicholaus Brosowsky"
date: "4/8/2019"
output:
  html_document:
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(dplyr)
library(mixtools)
```

## Simplified Expectation Maximization example

1. estimate the probabilities (expectation)
2. estimate the parameters (maximization)
3. repeat

```{r}
expectation <- function(data, means, proportions, sds){
  # Estimate the probabilities
  exp_data <- data %>% 
    mutate(prob_from_cluster1 = proportions[1] * dnorm(x, mean = means[1], sd = sds[1]),
           prob_from_cluster2 = proportions[2] * dnorm(x, mean = means[2], sd = sds[2]),
           prob_cluster1 = prob_from_cluster1/(prob_from_cluster1 + prob_from_cluster2),
           prob_cluster2 = prob_from_cluster2/(prob_from_cluster1 + prob_from_cluster2)) %>% 
    select(x, prob_cluster1, prob_cluster2)
    
  # Return data with probabilities
  return(exp_data)
}

maximization <- function(data_with_probs){
  means_estimates <- data_with_probs %>%
    summarise(mean_1 = sum(x * prob_cluster1) / sum(prob_cluster1),
              mean_2 = sum(x * prob_cluster2) / sum(prob_cluster2)) %>% 
    as.numeric()
  props_estimates <- data_with_probs %>% 
    summarise(proportion_1 = mean(prob_cluster1),
              proportion_2 = 1 - proportion_1) %>% 
    as.numeric()
  list(means_estimates, props_estimates)   
}

fun_gaussian <- function(x, mean, proportion){
  proportion * dnorm(x, mean, sd = 10)
}
```

# Create two gaussian distributions

```{r}
gaussian_sample <- data.frame(x = c(rnorm(7000, 20, 10), rnorm(3000, 50, 10)))

gaussian_sample %>% 
  ggplot() + geom_histogram(aes(x = x, y = ..density..), binwidth = 3, colour = "black", 
                 fill = "white") +
  theme_bw()
```

# Fit mixture model using expectation maximization function

```{r}
means_init <- c(0, 100)
props_init <- c(0.5, 0.5)

# Iterative process
for(i in 1:100){
  new_values <- maximization(expectation(gaussian_sample, means_init, props_init, c(10, 10)))
  means_init <- new_values[[1]]
  props_init <- new_values[[2]]
  cat(c(i, means_init, props_init), "\n")
}


```


# Plot estimated parameters over data

```{r}
gaussian_sample %>% 
  ggplot() + geom_histogram(aes(x = x, y = ..density..), binwidth = 3, colour = "black", 
                 fill = "white") +
  stat_function(geom = "line", fun = fun_gaussian,
                args = list(mean = means_init[1], proportion = props_init[1]),
                colour = "red", lwd = 1.5) +
  stat_function(geom = "line", fun = fun_gaussian,
                args = list(mean = means_init[2], proportion = props_init[2]),
                colour = "blue", lwd = 1.5) +
  theme_bw()

```

# Gaussian mixture model example #2

## Create two gaussian distributions

```{r}
# generate some data
set.seed(1523444)

# settings for data simulation:
N_samples    <- 1000       # number of total samples
prob1        <- 0.3        # probability of sampling from dist. 1 
prob2        <- 1 - prob1  # probability of sampling from dist. 2

sample_dist <- data.frame(
  mean = c(100,150),
  sd    = c(10, 20)
)

# create N_samples of prob1 and prob2
whichD <- sample(c(1,2),N_samples, replace=TRUE, prob=c(prob1, prob2))

# replace with samples from distribution
dat <- sapply(whichD, function(x) 
  rnorm(1,sample_dist$mean[x],sample_dist$sd[x]))

glimpse(dat)

```
# Fit mixture model using expectation maximization function

```{r}
# simulation starts with some guessed parameters
mu1    <- mean(dat,1)*0.8  # guess that dist1 has a mean smaller than the total sample average 
mu2    <- mean(dat,1)*1.2  # guess that dist2 has a mean larger than the total sample average
sd1    <- sd(dat)          # guess that sd = sample sd
sd2    <- sd(dat)
est_prop  <- 0.5           # guess the proportion of samples belonging to dist1
old_prop  <- 0             # placeholder for previous estimation


# function needed for the estimation maximization
# analogous to the weighted.mean function
weighted.sd <- function(x,w,mu = mean(x)){
  wvar <- sum(w*(x-mu)^2)/ sum(w)
  return(sqrt(wvar))
}

# iterate until threshold is reached
while (abs(est_prop-old_prop)>.00001){
  
  old_prop <- est_prop
  
  # Estimate the probabilities (expectation)
  resp <- est_prop*dnorm(dat,mu2,sd2)/
    ((1-est_prop)*dnorm(dat,mu1,sd1) + est_prop*dnorm(dat,mu2,sd2))
  
  # Estimate the parameters (maximization)
  mu1 <- weighted.mean(dat,1-resp)
  mu2 <- weighted.mean(dat,resp)
  
  sd1 <- weighted.sd(dat,1-resp,mu1)
  sd2 <- weighted.sd(dat,resp,mu2)
  
  est_prop <- mean(resp)
  print(est_prop)
  
}

df <- data.frame(rt=dat)
```

# Plot estimated parameters over data

```{r}
ggplot(df, aes(x = rt)) + 
  geom_histogram(aes(y = ..density..), colour = "black", fill = "white", binwidth = 3) + 
  stat_function(fun = function(k) (1-est_prop)*dnorm(k,mu1,sd1),
                colour = "red", lwd = 1.5) +
  stat_function(fun = function(k) est_prop*dnorm(k,mu2,sd2),
                colour = "blue", lwd = 1.5) +
  xlab("RT (ms)") + ylab("Density") +
  theme_bw()

# mixtools for comparison
```

# Run same as above, but with the mixtools library

```{r}
plot_mix_comps <- function(x, mu, sigma, lam) {
  lam * dnorm(x, mu, sigma)
}


mixmdl <- normalmixEM( dat, mu = c(1,4),
                     sigma=c(sd(dat),sd(dat)))


data.frame(x = mixmdl$x) %>%
  ggplot() +
  geom_histogram(aes(x, ..density..), binwidth = 3, colour = "black", 
                 fill = "white") +
  stat_function(geom = "line", fun = plot_mix_comps,
                args = list(mixmdl$mu[1], mixmdl$sigma[1], lam = mixmdl$lambda[1]),
                colour = "red", lwd = 1.5) +
  stat_function(geom = "line", fun = plot_mix_comps,
                args = list(mixmdl$mu[2], mixmdl$sigma[2], lam = mixmdl$lambda[2]),
                colour = "blue", lwd = 1.5) +
  ylab("Density") +
  theme_bw()
```
