---
title: "poisson_crime"
output:
  html_document:
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(ggplot2)
library(flexmix)
library(tidyr)
```

## Model clusters of neighborhoods by crime rates

```{r}
crimes <- read.csv("data/CoC_crimes.csv")
crimes_matrix <- as.matrix(crimes[,-1])
```

## Fit mixture model using poisson distributions
-  There are 15 different categories of crimes
-  find best fitting set of clusters (1 to 12) for crime categories (each neighborhood provides an observation)
-  Determine which neighborhoods fall within each cluster

```{r}
set.seed(11234)
poisson_mix_model <- stepFlexmix(crimes_matrix ~ 1, 
                                 k = 1:12, 
                                 nrep = 20, 
                                 model = FLXMCmvpois(),
                                 control = list(tolerance = 1e-15, iter = 1000))

```

## Use BIC method to determine which model was best

```{r}
best_fit <- getModel(poisson_mix_model, which = "BIC")

prior(best_fit)

param_pmm <- data.frame(parameters(best_fit))

param_pmm <- param_pmm %>% mutate(Type = colnames(crimes_matrix))

head(param_pmm)
```

## Plot clusters by crime rates


```{r}
param_pmm %>% 
  gather(Components, Lambda, -Type) %>% 
  ggplot(aes(x = Type, y = Lambda, fill = Type)) + 
  geom_bar(stat = "identity") +
  facet_wrap(~ Components) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1), 
        legend.position = "none")
```

## Show which neighborhoods that fall within each cluster
- Clusters 2, 3, and 5 have the highest crime rates
- Clusters 6 and 1 have the lowest


```{r}

crimes_c <- crimes %>% 
  mutate(CLUSTER = factor(clusters(best_fit)))

crimes_c %>% 
  group_by(CLUSTER) %>% 
  mutate(NUMBER = row_number()) %>% 
  ggplot(aes(x = CLUSTER, y = NUMBER, col = CLUSTER)) + 
  geom_text(aes(label = COMMUNITY), size = 2.3)+
  theme_bw() +
  theme(legend.position="none") +
  labs(x = "Clustered Neighborhoods")
```

